---
title: "JHU CSSE COVID-19 Dataset analysis"
author: "cloud-erik"
date: "21 5 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Import

First importing the COVID-19 Dataset as CSV file from JHU Github <https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data> and teh population and population density from UN <https://population.un.org/wpp/Download/Standard/CSV/>.

```{r import_data, echo=TRUE}
url_confirmed <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
url_death <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
url_recovered <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_recovered_global.csv"

url_population <- "https://population.un.org/wpp/Download/Files/1_Indicators%20(Standard)/CSV_FILES/WPP2019_TotalPopulationBySex.csv"

#check.names has to be FALSE because otherwise "/" will be replaced in Header names
covid_confirmed <- read.csv(url_confirmed, header = TRUE, check.names = FALSE)
covid_death <- read.csv(url_death, header = TRUE, check.names = FALSE)
covid_recovered <- read.csv(url_recovered, header = TRUE, check.names = FALSE)
population <- read.csv(url_population, header = TRUE, check.names = FALSE)

```

Clean the date and join in into one table. "transpose" with pivot_loger the dates into rows, drop the lat and long columns, make date a date format, country a valid name and drop the Province by sum it by country.
Keep in population only data from last year and current variant "2".

```{r cleaning, echo=TRUE}
library(tidyverse)
library(lubridate)

covid_confirmed <- covid_confirmed %>%
  pivot_longer(cols = -c("Province/State", "Country/Region", "Lat", "Long"),
               names_to = "date", values_to = "confirmed") %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long)) %>%
  rename(Country = "Country/Region") %>%
  group_by(Country, date) %>% summarise(confirmed = sum(confirmed))

covid_death <- covid_death %>%
  pivot_longer(cols = -c("Province/State", "Country/Region", "Lat", "Long"),
               names_to = "date", values_to = "death") %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long)) %>%
  rename(Country = "Country/Region") %>%
  group_by(Country, date) %>% summarise(death = sum(death))

covid_recovered <- covid_recovered %>%
  pivot_longer(cols = -c("Province/State", "Country/Region", "Lat", "Long"),
               names_to = "date", values_to = "recovered") %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long)) %>%
  rename(Country = "Country/Region") %>%
  group_by(Country, date) %>% summarise(recovered = sum(recovered))

# Join together in covid_data
covid_data <- merge.data.frame(covid_confirmed, covid_death)
covid_data <- merge.data.frame(covid_data, covid_recovered)

# rename Countries to be able to merge the data later together
# rename "US" in JHU Data with correct value "USA".
covid_data[["Country"]][covid_data[["Country"]]=="US"] = "USA"

#rename "Czechia" with "Czech Republic"
covid_data[["Country"]][covid_data[["Country"]]=="Czechia"] = "Czech Republic"

#rename "United Kingdom" with "UK"
covid_data[["Country"]][covid_data[["Country"]]=="United Kingdom"] = "UK"

summary(covid_data)

population <- population %>%
  filter(VarID == "2", Time == year(Sys.Date())-1) %>%
  select(-c(LocID, VarID, Variant, Time, MidPeriod, PopMale, PopFemale)) %>%
  rename(Country = "Location")

# rename Countries to be able to merge the data later together
#rename "United Kingdom" with "UK"
population[["Country"]][population[["Country"]]=="United Kingdom"] = "UK"
#rename "United States of America" with "USA"
population[["Country"]][population[["Country"]]=="United States of America"] = "USA"
#rename "Russian Federation" with "Russia"
population[["Country"]][population[["Country"]]=="Russian Federation"] = "Russia"
#rename "Iran (Islamic Republic of)" with "Iran"
population[["Country"]][population[["Country"]]=="Iran (Islamic Republic of)"] = "Iran"
#rename "Czech Republic" with "Czechia"
population[["Country"]][population[["Country"]]=="Czech Republic"] = "Czechia"
#rename "Brunei Darussalam" with "Brunei"
population[["Country"]][population[["Country"]]=="Brunei Darussalam"] = "Brunei"
#rename "Bolivia (Plurinational State of)" with "Bolivia"
population[["Country"]][population[["Country"]]=="Bolivia (Plurinational State of)"] = "Bolivia"
#rename "Côte d'Ivoire" with "Cote d'Ivoire"
population[["Country"]][population[["Country"]]=="Côte d'Ivoire"] = "Cote d'Ivoire"
#rename "Syrian Arab Republic" with "Syria"
population[["Country"]][population[["Country"]]=="Syrian Arab Republic"] = "Syria"
#rename "Venezuela (Bolivarian Republic of)" with "Venezuela"
population[["Country"]][population[["Country"]]=="Venezuela (Bolivarian Republic of)"] = "Venezuela"
#rename "Viet Nam" with "Vietnam"
population[["Country"]][population[["Country"]]=="Viet Nam"] = "Vietnam"
#rename "Viet Nam" with "Vietnam"
population[["Country"]][population[["Country"]]=="Viet Nam"] = "Vietnam"

covid_data <- merge.data.frame(covid_data, population, all.covid_data = TRUE)
```

First plot a timeline

```{r timeline, echo=TRUE}
library(timetk)

p <- ggplot(covid_data, aes(x = date)) +
  geom_line(color="darkorchid4", size=1, alpha=0.5, linetype=1, aes(y = confirmed)) +
  geom_line(color="yellow", size=1, alpha=0.5, linetype=1, aes(y = recovered)) +
  geom_line(color="blue", size=1, alpha=0.5, linetype=1, aes(y = death)) +
  labs(x="Date") +
  ggtitle("Covid19 over time")
p
```

Plot a map with confirmed cases

```{r map_confirmed, echo=TRUE}
library(ggmap)
library(ggplot2)

library(PBSmapping)
library(maps)
library(mapproj)

# create new data frame with sum per region (Country)
covid_region <- covid_data %>%
  rename(region = "Country") %>%
  group_by(region) %>% summarise(confirmed = max(confirmed),
                                  death = max(death),
                                  recovered = max(recovered),
                                  PopTotal = max(PopTotal),
                                  PopDensity = max(PopDensity)) %>%
  mutate(conf_per_pop = confirmed / PopTotal,
         death_per_pop = death / PopTotal,
         rec_per_pop = recovered / PopTotal)

myLocation<-c(-180, -60, 179, 85) # World
nMap <- get_stamenmap(bbox=myLocation, maptype="toner-lite", zoom=2)

#get country polygon data
mapdata <- map_data("world")
mapdata <- left_join(mapdata, covid_region, by="region")

#get bounding box for map
bb<-attr(nMap, "bb");
ylim<-c(bb$ll.lat, bb$ur.lat)
xlim<-c(bb$ll.lon, bb$ur.lon)

#clip polygons to map
colnames(mapdata)[1:6] <- c("X","Y","PID","POS","region","subregion")
mapdata<-clipPolys(mapdata, xlim=xlim, ylim=ylim, keepExtra=TRUE)

#plot map overlay
ggmap(nMap)+coord_map(xlim=xlim,ylim=ylim) +
    geom_polygon(data=mapdata, aes(x=X, y=Y, group=PID, fill=conf_per_pop), alpha=0.9) +  
    ggthemes::theme_map() +
    ggtitle("Confirmed Covid cases per 1000 population") +
    scale_fill_gradient(low = "yellow", high = "red", na.value = NA)


```

Plot a map with death cases

```{r map_death, echo=TRUE}

#plot map overlay
ggmap(nMap)+coord_map(xlim=xlim,ylim=ylim) +
    geom_polygon(data=mapdata, aes(x=X, y=Y, group=PID, fill=death_per_pop), alpha=0.9) +
    ggthemes::theme_map() +
    ggtitle("Death Covid cases per 1000 population") +
    scale_fill_gradient(low = "grey", high = "black", na.value = NA)

```

Plot a map with recovered cases

```{r map_recovered, echo=TRUE}

#plot map overlay
ggmap(nMap)+coord_map(xlim=xlim,ylim=ylim) +
    geom_polygon(data=mapdata, aes(x=X, y=Y, group=PID, fill=rec_per_pop), alpha=0.9) +
    ggthemes::theme_map() +
    ggtitle("Recovered Covid cases per 1000 population") +
    scale_fill_gradient(low = "yellow", high = "green", na.value = NA)


```

# Bias
There is plenty room of bias within the covid 19 data. First its collected all over the world from different organizations which could lead to offset and errors. Also the figures strongly depend on the infrastructure and the willing within a specific region to collect the data. e.g. there must be sufficient test capacity available and the people must be willing to be tested. Therefore its not reliable to compare e.g. figures from highly developed countries with huge testing capacity with countries where almost no health care system is in place.
In some areas also political effects influence the data like wars or if no data is shared.
And of course the values are absolute and have to be weighted or set in correlation of the population of the country. What was done by merging population data from UN into the dataset.

```{r esssionifo, echo=TRUE}
sessionInfo()
```
