---
title: "NYPD Shooting Incident Data"
author: "cloud-erik"
date: "20 5 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Import

First importing the "NYPD Shooting Incident Data (Historic)" Dataset as CSV file from <https://catalog.data.gov> and create a summary

```{r import_data, echo=TRUE}
url_source <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
shootings <- read.csv(url_source)
summary(shootings)
```

# cleaning up the data

* loading lubridate and tidyverse libary
* convert date and time column to one datetime
* drop not needed columns X_COORD_CD, Y_COORD_CD, Lon_Lat, PRECINCT, JURISDICTION_CODE, LOCATION_DESC, BORO and keep only one location (Latitude, Longitude)

After cleaning the data create again a summary

```{r clean_data, echo=TRUE}
library(lubridate)
library(tidyverse)
shootings <- shootings %>%
  unite(OCCUR_DATE, OCCUR_DATE:OCCUR_TIME) %>%
  mutate(OCCUR_DATE = mdy_hms(OCCUR_DATE)) %>%
  select(-c(INCIDENT_KEY, BORO, X_COORD_CD, Y_COORD_CD, Lon_Lat, PRECINCT,
    JURISDICTION_CODE, LOCATION_DESC)) %>%  
  rename(Murder = "STATISTICAL_MURDER_FLAG")
summary(shootings)
```

# Visulize

Because the shootings are spatial data I decided to do first a visualization on a map to get a better overview if there are any clusters.

first sort the data frame with murder at the tail to make sure that murder are not overprinted.
```{r sorting, echo=TRUE}
shootings <- shootings[order(shootings$Murder),]
```

## Map with shootings in blue and murder in red

First I plot all shootings on a stamen map of New York, shootings without murder in blue and with murder in red.
```{r map1, echo=TRUE}
library(ggmap)
library(ggplot2)
myLocation<-c(-74.27, 40.5, -73.7, 40.92) # New York
myMap <- get_stamenmap(bbox=myLocation, maptype="toner-lite", crop=TRUE)
ggmap(myMap)+
geom_point(aes(x=Longitude, y=Latitude), data=shootings, alpha=0.5, 
           color=ifelse(shootings$Murder=="true", "red", "blue"), size = 1)
```
## Heatmap wihh shootings in blue and murder in red

To make cluster more visual I visualize same data again with a heat-map starting from green up to shootings without murder in blue or with murder in red.

```{r map2, echo=TRUE}
ggmap(myMap)+
stat_density2d(aes(x=Longitude, y=Latitude,fill=..level.., alpha=..level..),
               data=shootings, geom="polygon")+
scale_fill_gradient(low = "green", high = ifelse(shootings$Murder=="true", "red", "blue"))+
theme(axis.ticks = element_blank(),
axis.text = element_blank(),
legend.position="none")
```
So it look that there happens significant more shootings and murder in the area of Brooklyn and Bronx than in the other areas.

# Timeline

To see if there is any trend over time I plotted data on a time-line aggregated by week.
```{r timeline, echo=TRUE}
library(timetk)
shootings_byweek <- summarise_by_time(shootings, .date_var = OCCUR_DATE,
                                      .by = "week", value = n())
p <- ggplot(shootings_byweek, aes(x = OCCUR_DATE, y = value)) +
geom_line(color="#69b3a2", size=1, alpha=0.9, linetype=1) +
geom_point()
p
```
It looks like these are some seasonal variances over the time and there is a peak of shootings in 2020.
Take only Data from one four Years to show the seasonal variance.
```{r timeline2014-2017, echo=TRUE}
library(timetk)
shootings_byweek2015 <- filter_by_time(shootings_byweek, .date_var = OCCUR_DATE,
                                       .start_date = "2014-01-01", .end_date = "2017-12-31")
p <- ggplot(shootings_byweek2015, aes(x = OCCUR_DATE, y = value)) +
geom_line(color="#69b3a2", size=1, alpha=0.9, linetype=1) +
geom_point()
p
```
So it look that there happens more shootings during summer and lesser during winter time.

# Bias
Because shootings are surpassing events I expect that there is no great bias like in other criminal statistics with minor crime that could be e.g. somehow correlated with the presence of police. But I assume that shootings will be recognized always with or without police and so all shootings should be part of the official statistics.
Also the correlation between murder and shootings without murder shows that there seems minor bias in the data.

The spatial data should take into account the density of population in more detailed analysis.

```{r esssionifo, echo=TRUE}
sessionInfo()
```